\chapter{QDOAS Overview} \label{cha:intro}
\section{Introduction}
The experience of the \gls{bira} in the development and improvement of algorithms for the retrieval of trace gas concentrations goes back to the early 1990s, with atmospheric research activities using ground-based UV-Visible spectrometers aiming at the long-term monitoring of minor components involved in the catalytic destruction of the ozone layer or in anthropogenic pollution.

WinDOAS, the first program developed at \gls{bira} in 1997, knew a success story due to a friendly user interface completed with some po\-wer\-ful \glslink{doas}{DOAS} tools.  This program, extensively validated through different campaigns, has been used worldwide and for many different DOAS applications (mainly for ground-based and satellite applications).

QDOAS is a cross-platform implementation of WinDOAS: the software is portable to Windows and Unix-based operating systems whereas WinDOAS was designed only for Windows. The user interface and the engine of QDOAS are similar to those of its predecessor.  WinDOAS is no longer supported.

QDOAS has been developed in collaboration with
\href{http://www.stcorp.nl/}{S[\&]T}, a Dutch company well-known for
the development of cross-platform products and software tools for the
processing of satellite measurements (\glslink{beat}{BEAT}, VISAN,
\ldots). The graphical user interface is built on the Open-Source
version of the Qt toolkit, a cross-platform application framework,
and Qwt libraries.

QDOAS is free software distributed under the terms of the GNU General Public License; it is open source and the code is available on request by contacting the \href{mailto:caroline.fayt@aeronomie.be,thomas.danckaert@aeronomie.be}{authors}.
This document describes the QDOAS user interface and dialog boxes for configuring the software.  It completes the online help provided with the package.  The algorithms that are used in the software are also summarized.  The document assumes that users  already have a minimum of experience with \gls{doas} applications and retrieval.

\section{Main QDOAS features}
\subsection{General features}
\begin{qdoasitem}
\item The main components of the \gls{gui} are organized in multi-page panels with a fixed arrangement and tab-switched access to the different pages;
\item The application is based on a tree structure;
\item All the spectral windows are processed in one shot and there is a tab-switched access between the different fitting windows;
\item Large amount of files can be processed in one shot;
\item Support different spectra file formats (see section \ref{sec:intro-formats}) and specificities for satellites, ground-based and airborne measurements are accounted for;
\item On line help in HTML format
\end{qdoasitem}
\subsection{Plot}
\begin{qdoasitem}
\item Visualization of spectra and the results in different tab pages;
\item Possibility to set plot contour and style;
\item Interactive plot mode (zooming, overlay of an existing ASCII file, possibility to fix the scaling of the plot, \ldots): activated by right clicking the title of the plot ;
\item Export of plot in different portable image formats (png, jpg)
\end{qdoasitem}
\subsection{Analysis}
\begin{qdoasitem}
\item DOAS/intensity fitting modes;
\item shift/stretch (in nm) fully configurable for any spectral item (cross section or spectrum);
\item automatic reference selection from the input spectra file taking account for specificities for satellites or MAXDOAS measurements (scans);
\item possibility to constrain the slant column density of a molecule to the value found in a previous window;
\item possibility to filter spectra and cross sections before analysis (supported filters include Kaiser, gaussian, boxcar, Savitsky Golay…);
\item possibility to define gaps or automatically remove spikes within fitting intervals (e.g. to eliminate bad pixels);
\item possibility to fit an instrumental offset;
\item possibility to define several configurations of spectral windows under a project;
\end{qdoasitem}
\subsection{Calibration And Slit Function Characterization}
\begin{qdoasitem}
\item wavelength calibration and instrumental slit function characterization using a \gls{nlls} fitting approach where measured intensities are fitted to a high resolution solar spectrum degraded to the resolution of the instrument. The fitting method (DOAS or intensity fitting) can be different from the method used in the analysis;
\item possibility to correct for atmospheric absorption and Ring effect;
\item supports different analytical line shapes, as described in section \ref{sec:wavel-calibr}, page \pageref{sec:wavel-calibr};
\item allows approximating the initial shift of the NLLS procedure by calculating the best correlation between the spectrum to calibrate and the solar spectrum (cfr the \textbf{calculation of the preshift}, page \pageref{sec:projects_calibration_preshift} );
\end{qdoasitem}

\subsection{Cross sections handling}
\begin{qdoasitem}
\item possibility to calculate differential absorption cross sections (by orthogonalization or high-pass filtering);
\item possibility to multiply cross sections with wavelength dependent \glspl{amf};
\item possibility to fix the column density of any selected species;
\item possibility to convolve cross sections in real time using a user defined slit function or the information on calibration and slit function provided by the wavelength calibration procedure;
\item possibility to handle differences in resolutions between measured and control spectra;
\end{qdoasitem}

\subsection{Output}
The output is fully configurable. Analysis results and various data related to the measurements can be saved to ASCII, HDF-EOS5 or netCDF files.

\subsection{Tools}

The package includes five independent executable or modules~:

  \begin{qdoasitem}
  \item qdoas~: the main user interface;
  \item convolution~: the convolution/filtering tool;
\begin{itemize} \itemsep -2pt
\item standard and \I0-corrected convolutions are supported;
\item possibility to create an effective slit function taking into account the (finite) resolution of the source spectrum (using a FT deconvolution method);
\item asymmetric line shapes, wavelength dependent slit functions;
\end{itemize}

  \item ring~: The ring tool calculates Ring effect cross sections (Rotational Raman Scattering approach);
  \item usamp~: This tool generates undersampling cross sections;
  \item doas\_cl~: a powerful command line tool that applies on qdoas, convolution, ring and usamp configuration files (XML format).
  \end{qdoasitem}
Convolution, ring and undersampling tools can be also called from the QDOAS user interface.

\section{Supported spectra file formats} \label{sec:intro-formats}
\subsection{Satellite Measurements}
Spectra measured by the following satellite instruments can be processed by QDOAS~:

\subsubsection{GOME (ERS-2)}

QDOAS supports the original ASCII file format but it is recommended to use the more suitable binary file format created by a modified version of the \glslink{gdp}{GDP} implemented at \gls{bira} for the automatic selection of the reference spectrum.  Contact the authors for further information about this format.

\subsubsection{SCIAMACHY (ENVISAT)}

Routines to read the \glslink{sciamachy}{SCIAMACHY} PDS file format have been kindly provided by IFE/IUP University of Bremen.  SCIAMACHY Level 1C versions up to 7.04 (the SCIAL1C utility converts files from L1 to L1C)

\subsubsection{GOME2 (MetOp)}

QDOAS uses the \glslink{coda}{CODA} library to read spectra from GOME2.  The
package can be downloaded from the S[\&]T web site. Before using QDOAS
on GOME2 spectra, CODA should be installed and the
\texttt{CODA\_DEFINITION} environment variable should be defined.

QDOAS supports GOME2 Level 1B, PPF 5.0 product format version 12.0 and
below.

\subsubsection{OMI (AURA)}

\glslink{omi}{OMI} spectra are read using the
\glslink{hdfeos}{HDF-EOS2} library, which is based on
\glslink{hdf}{HDF4}.

\subsection{Ground-based Measurements}

The most popular file formats supported by QDOAS for ground-based measurements are~:

\subsubsection{SAOZ}

Système d'Analyse par Observation Zénitale developed by the aeronomy lab of the CNRS (JP Pommereau, F Goutail,...), France, and largely used in the \glslink{ndacc}{NDACC} network for stratospheric total ozone and \NO2 monitoring.   Both PCD/NMOS 512 and EFM 1024 (SAM) formats are supported.

\subsubsection{MFC format, DOASIS} \label{sec:MFC_DOASIS}

Both binary and std (ASCII) formats are produced by the well-known \glslink{doasis}{DOASIS} program designed by the atmospheric research group at IUP Heidelberg, Germany and widely used in the DOAS community.  It can be obtained from the following URL:

\url{https://doasis.iup.uni-heidelberg.de/bugtracker/projects/doasis}

This format can also be produced by other programs or converted from other formats, e.g.\ the Spectra Suite Ocean Optics software. QDOAS requires the size of the detector.
The MFC binary format generated by DOASIS is supported by QDOAS since version 2.111.  The specificity of this format is that spectra are saved in individual files that are not easy to manage mainly for the automatic reference selection (zenith-sky spectrum measured at the minimum SZA of the day or zenith-sky of the scan for MAXDOAS measurements).
To browse and process them, it is recommended to insert folders instead of individual files.  Browsing records requires the use of the file selection combobox in the toolbar (cfr page \pageref{sec:interface_toolbar} ) as one file contains only one record in this format.  Nevertheless, the
automatic reference selection is automatically performed by browsing all the files of the current directory.  This could make the analysis heavier and a conversion to the BIRA-IASB MFC Binary format is recommended (using the \texttt{MFC\_Std2Bin} converter available on the FTP server with QDOAS)

\subsubsection{MFC binary, BIRA-IASB} \label{sec:MFC_bin}

Because spectra are saved in individual files with automatic numbering, MFC STD format is not very suitable to browse spectra and for efficient automatic reference selection (zenith of the scan for MAXDOAS data).  For this reason, BIRA-IASB has developed its own MFC binary format
in which all spectra of a same day including dark currents and offsets measured the night are saved in a unique file.  When loading a file, measurements are automatically corrected by the averaged dark currents and offsets.  The \texttt{MFC\_Std2Bin} utility that converts
MFC STD files in the MFC BIRA-IASB binary format is available with the QDOAS package since version 2.107.  Syntax to use from command line is

\begin{quote}
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2}]
MFC_Std2Bin <config file> <input file or path> <output path>

where :

  config file        : is the name of a file with some options
  input file or path : the name of the MFC file or path to process
  output path        : the path where to save the data

It is recommended to apply on dark current/offset files or paths
before spectra.  Daily output files are automatically generated
using the dates of measurement.  Yearly directories are created
in the output path.

The config file is an ASCII file including lines starting with the
following keywords :

   date format=...      : the format of the date, for example YYYYMMDD
   spectrum=...         : the string identifying spectra measurements
   offset=...           : the string identifying offset measurements
   dark=...             : the string identifying dark current measurements
   output prefix=...    : the prefix of output file name (the date in
                          YYYYMMDD format will follow)
   output extension=... : the file extension to complete the output file name

The program will use the strings identifying spectra, offset and dark
current to attribute a measurement type number to spectra :

   1 for off axis
   3 for zenith sky
   4 for dark currents
   8 for offsets

Example of config file :

   date format=MMDDYYYY
   spectrum=Measurement Spectrum
   offset=Offset Spectrum
   dark=Dark Current Spectrum
   output prefix=BX
   output extension=bin

All files of 07/01/2014 will be saved in <output path>/2014/BX_20140107.bin
\end{Verbatim}
\end{quote}

\subsubsection{MKZY PAK}

developed at the Chalmers University of Technology, Goteborg, Sweden (MANNE Kihlman and ZHANG Yan) and used in the NOVAC network.

\subsubsection{OCEAN OPTICS}

very simple ASCII file format developed by the spectrometers manufacturer.

\subsection{The ASCII File Format}
Other formats are specific to the different institutes that developed them.  If a format is not supported by QDOAS, it should be possible to convert the files to ASCII. When the ASCII format is selected, spectra can be provided in the file one record per line \textbf{(line format)} or one spectral value per line \textbf{(column format)} or several spectra in columns (from version 2.107). According to the checked flags in the \textbf{Read from file} group box, the following information are expected strictly in the given order~:
\begin{itemize} \itemsep -6pt
\item Solar Zenith Angle
\item Azimuth Viewing Angle
\item Elevation Viewing Angle
\item Date in the DD/MM/YYYY (day/month/ year) format
\item Fractional time
\end{itemize}

\textbf{In the column format, angles had to be given on the same line until version 2.106.
From version 2.107, the column format accepts also matrices of spectra and angles have to be given in separate lines.}  The file can contain
several matrices of spectra but the number of columns has to be the same for all data sets.

\begin{quote}
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2}]
Example of ASCII file in column format :

* data set with 4 individual spectra
* line 1 : solar zenith angle
* line 2 : elevation angle (azimuth viewing angle not provided here)
* line 3 : date
* line 4 : fractional time
* remaining lines : spectra values
   9.5007600e+01   9.3970000e+01   9.3260500e+01   9.3067300e+01
   8.5000000e+01   8.5000000e+01   8.5000000e+01   8.5000000e+01
   01/07/2012      01/07/2012      01/07/2012      01/07/2012
   2.9940300e+00   3.1491700e+00   3.2526400e+00   3.2804200e+00
   0.0000000e+00   0.0000000e+00   0.0000000e+00   0.0000000e+00
  -1.8073600e+00  -5.0432300e+00  -6.7440700e+00  -4.7440500e+00
  -1.4793200e+00  -3.5119500e+00  -2.1264100e+00  -9.1264000e+00
   ...             ...             ...             ...
\end{Verbatim}
\end{quote}

With column format, a wavelength calibration can be provided with spectra.  It should
be given in first column of each data set.

Other examples of ASCII files supported by QDOAS are given in appendix \textbf{\ref{cha:input-files-format}}.

\section{System requirements and installation}

\subsection{System Requirements}

The \gls{gui} is built on the Open-Source version of the Qt
toolkit. As a result, QDOAS is portable to Windows, Linux and macOS,
and the user interface is effectively the same on all platforms. The
software is free and distributed under the GNU GPL version 2.0 (see
\url{http://www.gnu.org/licenses/gpl-2.0.html} for further details).

To obtain the software, get support and receive notificatations of new
releases, please register on the following page:

\url{http://uv-vis.aeronomie.be/software/QDOAS/QDOAS_Register.php}

\subsection{Install QDOAS on Windows}

Windows users can download the executables and all needed dynamic libraries from our FTP server (address provided after registration).
\subsubsection{GOME2}
For GOME2 applications, the \texttt{CODA\_DEFINITION} environment
variable must be defined.  This variable should point to the file
system location where the .codadef file that describes the file format
of the GOME2 input data can be found (currently,
EPS-20150324.codadef).  In windows, it can be set via

Control Panel $\rightarrow$ System $\rightarrow$ Advanced $\rightarrow$ Environment variables

\subsection{Install QDOAS on Linux}
Users who wish to use QDOAS on Linux can use the binary package
provided on our FTP server.  This package should run on almost all
recent Linux versions.

Users can also choose to compile the program from the source code.
For systems where the \href{https://www.gnu.org/software/guix}{Guix}
package manager is available, a package description is provided in the
file \texttt{qdoas.scm}, found in the top directory of the QDOAS
source package.  After unpacking the source, QDOAS can be installed
with the following command:
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
guix package --install-from-file=/path/to/source/qdoas.scm
\end{Verbatim}
For other systems, this section contains instructions to compile QDOAS
and some of the libraries it depends on.

Users without administrator privileges usually have to install QDOAS
and any extra required libraries in their home directory.  In the
following instructions, we assume that the user wants to install qdoas
and any libraries that might need to be installed in the directory,
\texttt{/home/username}.  Executable files will be installed in
\texttt{/home/username/bin}, libaries and headers in
\texttt{/home/username/lib} and \texttt{/home/username/include}.  When
you want to install the software in a different directory, you will
need to replace \texttt{/home/username} by the directory of choice in
the following instructions.

In order that the operating system can find the QDOAS executable
files, the \texttt{bin} subdirectory must be added to the
\texttt{\$PATH} system variable.  To do this, the user will add the
following entry in to their shell (Bash) profile (typically the file
\texttt{/home/username/.profile}), or execute this command directly
from the shell\footnote{When you execute this command from the shell,
  the changes are lost at the end of your session, and you will have
  repeat the command every time you start a new session.  Add the
  command to your shell profile to have it executed automatically at
  the start of every session.}:
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
export PATH=/home/username/bin:$PATH
\end{Verbatim}
%$

\subsubsection{C and C++ compiler}
QDOAS requires gcc and g++ versions 4.7 or higher, or recent versions of the Clang compiler.

\subsubsection{Qt}
Qdoas uses the Qt libraries.  It can be compiled with Qt5 or, at your
choice, Qt4.

Qt development packages are available on most Linux distributions, as
well as on macOS (e.g.\ via Homebrew or MacPorts).  Alternatively,
you can install Qt yourself from \url{https://www.qt.io/download} and
follow the installation instructions.

\subsubsection{Qwt}
Either install a Qwt development package for your OS, or install the
Qwt library using the source code, available at
\url{http://sourceforge.net/projects/qwt}.

To install Qwt manually, download the latest source package from the
Qwt project website and unpack it in a temporary directory. The
current QDOAS version requires Qwt version 6.1.2 or higher.

Open a shell and enter the Qwt source directory, edit the
qwtconfig.pri and change the variables according to your installation.
A sysadmin can set the \texttt{QWT\_PREFIX\_INSTALL} to, for example,
\texttt{/usr/local}, while a regular user might set it to
\texttt{/home/username}.

Once the changes are made accordingly to your installation, run the
command \textbf{qmake qwt.pro} followed by \textbf{make} and finally
\textbf{make install}. This will build the Qwt libraries and install
the development package (headers, documentation and runtime
libraries).

\subsubsection{GSL}
QDOAS uses numerical routines from the GNU Scientific Library.  You can download this library from \url{https://www.gnu.org/software/gsl}.

From the unpacked source directory, run 
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
$ ./configure --prefix=/home/username
\end{Verbatim}

And continue with \textbf{make} and \textbf{make install}.

\subsubsection{CODA}\label{sec:coda}
QDOAS uses the \gls{coda} library developed by S[\&]T to read GOME-2
data.  The source package is available at
\url{http://www.stcorp.nl/coda}.

CODA can be installed with the usual ``configure/make/make install''
procedure.  From the directory where you have downloaded the CODA
library files, enter

\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
$ ./configure --prefix=/home/username
\end{Verbatim}
%$

This configures the install script to use the directory structure we
described earlier.  To complete the installation, type \texttt{make},
which will compile the library files, and finally \texttt{make
  install}, which will copy the compiled files, headers and other
resources to the appropriate subdirectories in \texttt{/home/username}.

If you want to process GOME-2 data using QDOAS, you need to obtain the
product format descriptions for the GOME-2 data as well.  Download the
files for ``EUMETSAT EPS GOME-2 and IASI products'' from
\url{http://stcorp.nl/coda/registration/download}, save them in a
location such as \texttt{/home/username/share/coda/definitions}, and
define the environment variable \texttt{CODA\_DEFINITION} so it
contains the location of the .codadef files.  For example:

\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2}]
In bash:
   export CODA_DEFINITION="/home/username/share/coda/definitions"
In c-shell:
   setenv CODA_DEFINITION "/home/username/share/coda/definitions"
\end{Verbatim}

Again, it is practical to add these commands to your shell profile.

\subsubsection{HDF-EOS2}
QDOAS uses the HDF-EOS2 library to read OMI spectra.  The HDF-EOS2
library itself relies on the HDF4 library.  If these libraries are not
yet available on your system, you can install them using your Linux
distribution's package management system, or install them from the
source code.

If you want to install the libraries from source, HDF4 must be
installed first.  The HDF4 source code package is available from
\url{http://www.hdfgroup.org/ftp/HDF/HDF_Current/src}.  You can
install it using the usual ``configure/make/make install'' procedure.
To avoid conflicts with the netCDF library, you should disable the
netCDF interface of the HDF4 library using the configure option
\texttt{--disable-netcdf}. To avoid a conflict with the HDF5 library,
we have found it practical to install the HDF4 header files in a
separate subdirectory using the option \texttt{--includedir}.  The
\texttt{configure} command which achieves this is:

\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2}]
$ ./configure --prefix=/home/username --disable-netcdf \
    --includedir=/home/username/include/hdf4
\end{Verbatim}
%$

Now, use \texttt{make} and \texttt{make install} to complete the
installation of HDF4.

Once you have installed the HDF4 library, download the source for
HDF-EOS2 from
\url{ftp://edhs1.gsfc.nasa.gov/edhs/hdfeos/latest_release}.  The
library comes with an installation script, but we recommend installing
it with the same ``configure/make/make install'' procedure, which is
also available.  If you installed HDF4 according to the previous
instructions, the following command should allow you to install
HDF-EOS2:
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2}]
$ ./configure --prefix=/home/username CC=/home/username/bin/h4cc \
    --enable-install-include
\end{Verbatim}
%$

When \texttt{configure} has finished its job, run \texttt{make} and \texttt{make install}.

\subsubsection{HDF5}
QDOAS can write output data in the HDF-EOS5 and netCDF-4 formats.
Both formats depend on the HDF5 library.  Again, if these libraries
are not installed on your system yet, you can either install the
appropriate packages for your Linux distribution, or compile the
libraries from source yourself.

Source code packages for HDF5 are available from
\url{http://www.hdfgroup.org/ftp/HDF5/current/src}.  The installation
follows the standard ``configure/make/make install'' procedure.  When
running configure, specify the \texttt{--prefix} as before:

\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
$ ./configure --prefix=/home/username --enable-install-include
\end{Verbatim}

Again, run \texttt{make} and \texttt{make install} to complete the
installation.

\subsubsection{netCDF}
The netCDF source code is available from
\url{http://www.unidata.ucar.edu/software/netcdf}.  QDOAS uses the
netCDF-4 format, which depends on the HDF5 library we've just
installed.  To ensure that the netCDF installation scripts can find
your HDF5 library, use the following \texttt{configure} command:

\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
$ CPPFLAGS=-I/home/username/include LDFLAGS=-L/home/username/lib \
    ./configure --prefix=/home/username
\end{Verbatim}

followed by \texttt{make} and \texttt{make install}.

\subsubsection{HDF-EOS5}
After installing HDF5, you can obtain the HDF-EOS5 source
code\footnote{The current version 1.15 of the HDF-EOS5 source code
  contains a bug which causes linking errors when the library is
  compiled without the SZIP library.  When installing HDF-EOS5 1.15
  from source, please appy the patch hdf-eos5\_1.15.patch, also
  available on our FTP server.} at
\url{ftp://edhs1.gsfc.nasa.gov/edhs/hdfeos5/latest_release}.  This
package can then be installed using configure and make.  Again, you
should set \texttt{--prefix=/home/username} to install the library at
the correct location.  Additionally, you need to add the option
\texttt{--with-hdf5=/home/username}, so the installation script can
find the HDF-5 libraries installed in the previous step, and
\texttt{--enable-install-include}, in order to copy the HDF-EOS5
header files to the directory \texttt{/home/username/include}, where
the Qdoas build system expects them:

\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
$ ./configure --prefix=/home/username --with-hdf5=/home/username \
    --enable-install-include
\end{Verbatim}
%$

Complete the installation of HDF-EOS5 by running \texttt{make} and
\texttt{make install}.

\subsubsection{QDOAS}
The QDOAS build configuration uses the Qt utility \textbf{qmake}.  The
configuration file is \textbf{Src/config.pri}, and some important
settings are made in the \texttt{unix} section:
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
unix {
  INCLUDEPATH += $$(HOME)/include
  QMAKE_LIBDIR += $$(HOME)/lib
  QMAKE_RPATHDIR += $$(HOME)/lib

  isEmpty(INSTALL_PREFIX) {
    INSTALL_PREFIX = $$(HOME)
  }

  INCLUDEPATH += $$INSTALL_PREFIX/include
  INCLUDEPATH += $$INSTALL_PREFIX/include/hdf4
  INCLUDEPATH += $$INSTALL_PREFIX/include/qwt
  QMAKE_LIBDIR += $$INSTALL_PREFIX/lib
}
\end{Verbatim}

This configuration assumes that QDOAS should use libraries installed
in the user's home directory (e.g. \texttt{/home/username}).  If you
followed the instructions until now, you not need to make any changes
to this file.  If some of the include files and/or libraries are
installed in other locations, you may have to add the include files'
and library files' directories to the variables \texttt{INCLUDEPATH}
and \texttt{QMAKE\_RPATHDIR}/\texttt{QMAKE\_LIBDIR} respectively.
Conversely, if QDOAS may not use libraries from the home directory,
remove the lines referring to \texttt{\$\$(HOME)}.

When you are satisfied with \texttt{config.pri}, go to the directory
\texttt{Src} contained in the QDOAS source distribution. Run the
command \textbf{qmake all.pro} followed by \textbf{make}. This will
create the executable files, \texttt{qdoas}, \texttt{doas\_cl},
\texttt{convolution}, \texttt{ring} and \texttt{usamp}.  The command
\texttt{make install} copies the executable files to the directory
\texttt{\$\$INSTALL\_PREFIX/bin}.

QDOAS uses the Qt resource system so that the executable only depends
on system Qt and Qwt libraries. The executable can therefore be safely
moved to another directory.  To rebuild QDOAS after an update of the
package, run the following commands from the \texttt{Src} directory:

\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
$ make distclean
$ qmake all.pro
$ make
\end{Verbatim}
%$

\subsection{QDOAS on macOS}
Users who wish to run QDOAS on a Mac system have to compile the program from the source code.  We suggest to use the MacPorts package manager, which can automatically install all required libraries, with the exception of CODA.  In the following instructions, we use \texttt{/Users/username} to refer to the user's home directory.

\begin{enumerate}
\item Install MacPorts: follow the installation instructions at \url{https://www.macports.org/install.php}.
\item Using MacPorts' \texttt{port} tool, install Qt5, Qwt, HDF-EOS2, HDF-EOS5, netCDF and GSL: open a terminal and enter
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
$ sudo port install qt5 hdfeos hdfeos5 netcdf gsl qwt61 +qt5
\end{Verbatim}
\item Install CODA: download and extract the CODA source code (see section \ref{sec:coda}), open a terminal, navigate to the location of the extracted source code, and enter
  \begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
$ ./configure --prefix=/opt/local
\end{Verbatim}
\item Extract the QDOAS source code archive.
\item Optionally: change the default install location.  By default, the QDOAS applications will be installed in \texttt{/Users/username/bin}.  If you want to install QDOAS in a different location, edit the file \texttt{Src/config.pri} from the QDOAS source bundle and look for the section 'macports'.
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
  macports {
  INSTALL_PREFIX = $$(HOME)
  INCLUDEPATH += [...]
\end{Verbatim}
  You can replace \texttt{\$\$(HOME)} by a directory of your choice,
  and save the file.
\item Open a terminal, navigate to the \texttt{Src} directory of the extracted QDOAS source code and enter the following commands:
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
$ /opt/local/libexec/qt5/bin/qmake CONFIG+=macports
$ make
$ make install
\end{Verbatim}
  In the second step, you can speed up the compilation process using
  make's parallel option, i.e.\ use \texttt{make -j8} to run up to 8
  jobs in parallel.
\end{enumerate}

You can now run the programs \texttt{qdoas}, \texttt{doas\_cl}, \texttt{convolution}, \texttt{ring} and \texttt{usamp} from the
terminal as follows:
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
$ /Users/username/bin/qdoas.app/Contents/MacOS/qdoas
$ /Users/username/bin/doas_cl.app/Contents/MacOS/doas_cl
$ ...
\end{Verbatim}
To avoid having to enter these long paths every time you want to run QDOAS, you can create aliases for the QDOAS applications in your shell profile.  For example, to create an alias \texttt{qdoas}, edit (or, if the file does not exist, create) the file \texttt{/Users/username/.bash\_profile}, and add the following line:
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
alias qdoas="/Users/username/bin/qdoas.app/Contents/MacOS/qdoas"
\end{Verbatim}
In your next session you can start QDOAS from the terminal as follows:
\begin{Verbatim}[fontfamily=courier,fontsize=\relsize{-2},xleftmargin=5mm]
$ qdoas
\end{Verbatim}
\subsection{Online Help}
QDOAS also provices online help in the form of HTML pages.  The
content is similar to chapters \ref{cha:descr-algor},
\ref{cha:properties} and \ref{cha:tools} of this document.  The HTML
files have to be copied on your disk (usually a subfolder Help of the
directory where executables are installed on Windows systems; a
dedicated subfolder of your system share directory on Linux systems).
The HTML pages are reachable from the Help button of the configuration
dialog boxes.  If the page can not be found, the user is is asked to
provide the location of the help's index.html file, after which the help system should work.

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "QDOAS_manual"
%%% TeX-PDF-mode: t
%%% End:
